From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-04-16 appendixB/appendixB.Rmd
nearly there

>>>>

\newpage

# Stochasticity in more complex models

The literature exploring how stochasticity plays out in more complex models, such as with age or stage structure, spatial structure, evolutionary dynamics and so forth is far more extensive than can be given justice here.  Without any suggestion of being a comprehensive review, the following table merely highlights just a few of the classic textbooks, papers, and reviews for any reader looking to explore further, as well as a handful of more recent papers merely to illustrate that these all remain active areas of research.  Addressing the interpretation, role, and consequences of stochasticity in each of these contexts would provide a more traditional introduction to the subject, such as a graduate course or seminar.   In the review and synthesis of the main text I have endeavored instead to deviate from this structure and constrain the focus to simpler models to better underscore the paradigms in which we do and might think about noise throughout the field.

Reference        |  Contexts          |  Description / Notes
-----------------|--------------------|--------------------------------------------------
@Cohen1979       | Age/stage structure         | Derives strong ergodic theorem for convergence to normalized age structure
@Tuljapurkar1989 | Age/stage structure         | Classic review of demography in random environments
@Tuljapurkar1997 | Age/stage structure         | Chapter on stochastic matrix models; from Tuljapurkar and Caswell's classic text on structured population modeling
@Caswell2009     | Age/stage structure         | Nice recent paper exploring individual stochasticity in age and stage structured populations, with examples from right whales and a prairie plant. 
@Metcalf2015     | Age/stage structure         | Excellent methods paper for using Integral Projection Models in stochastic population dynamics
@Vindenes2017b   | Age/stage structure         | 
@Dieckmann2000   | Spatial structure           | The book "The Geometry of Ecological Interactions" provides wide-ranging examples on approximations to spatially explicit dynamics, including stochastic context.
@Durrett1994     | Spatial structure & Indiv heterogeneity       | Classic paper on "Importance of Being Discrete and Spatial", includes nice demonstration of how spatially explicit stochastic models with discrete individuals can give qualitatively different behavior from their associated limiting reaction-diffusion (i.e. a deterministic, continuous PDE) models.
@Vindenes2015    | Indiv heterogeneity         | Very nice example of a recent paper incorporating individual heterogeneity into integral projection model methods for structured populations.
@Hart2016        | Indiv heterogeneity         | Another elegant recent example of incorporating individual heterogeneity, this time in a competition model, exploring impact on coexistence
@Tuljapurkar1980 | Coexistence                 | Proves convergence to the log-normal distribution for stochastic Leslie matrices
@Chesson1981     | Coexistence                 | Classic paper introducing what we now call the 'temporal storage effect' for coexistence of multiple species on the same resource through a varying (including but not limited to stochastic) environments. 
@Chesson1985     | Coexistence                 | Storage effect in time and space -- again, stochasticity can be a driver of the necessary variation across space as well as time.
@Schreiber2017   | Coexistence                 | Excellent overview of modern coexistence theory under both cases of demographic and environmental noise.  Includes several theorems showing how classic results (e.g. Chesson and others) can be extended to more general assumptions and also highlights some open questions and conjectures.  
@Roughgarden     | Colored noise               | Elegant simple model showing how auto-correlated environments differ from the dynamics predicted by classic white-noise approximations
@Ripa            | Colored noise               | Another simple model of colored noise in which extinction risk decreases with positively auto-correlated noise and decreases with negatively-auto-correlated noise
@Schreiber2010   | Colored noise               | Excellent illustration of how the impact of temporal auto-correlation in noise will depend on spatial heterogeneity and dispersal. (In particular, proves that a metapopulation in which expected fitness in every patch is less than 1 can still persist(!) given positive temporal auto-correlation in relative fitness and weak spatial correlation.)
@Lee2017         | Colored noise               | Recent numerical study which stands out from most other examples in this list by arguing that under a wide range of parameterizations they consider, ignoring auto-correlation has only a limited impact on estimating expected extinction times. 
@Paniw2018       | Colored noise               | Another recent numberical study demonstrating that autocorrelation most impacts stochastic population growth rates in populations with relatively fast life-histories.
@Coulson2001     | periodic/temporal           | Landmark paper demonstrating the importance of interactions between population structure / individual heterogeneity and environmental stochasticity using long-term data from sheep on St. Kilda.
@Bjornstad2001   | periodic/temporal           | Landmark paper highlighting the tension between 'deterministic' and 'stochastic' explanations for fluctuating populations.
@Keeling2001     | periodic/temporal           | Nice example of seasonally forced switching between attractors, as discussed in Stochastic Switching section of main text.
@Saether1997     | eco-evolutionary            | Great early review of environmental stochasticity in population dynamics emphasizing trait-based mechanisms and potential evolutionary consequences of that variation.
@Vindenes2015    | eco-evolutionary            | Incorporating individual heterogeneity (i.e. individual traits) also allows the authors to explore evolutionary consequences on this variation
@Schreiber2015   | eco-evolutionary            | Evolution of bet-hedging strategies in variable environments (a nice generalization of classic results from Gillespie 1973, 1974 through varying level of correlation within generations)
@Lenormand2009   | evolutionary                | Excellent recent review of stochasticity in evolutionary models

\pagebreak

### 2018-04-16 paper/paper.Rmd
nearly there

This distinction between a paradigms of noise the "nuisance" versus noise the "creator"
is necessarily a simplification across a spectrum of perspectives. These categories
should not be mistaken for either a sharp dichotomy nor a reference to a stictly
empirical-theoretical divide.  It is also essential to recognize that each paradigm
expands upon rather than rejects the previous notion of noise: the recognition that
noise can create novel phenomena does not mean that noise cannot also obscure the 
signal of some process of interest.  Likewise, seeking to use noise as a novel source
of information about underlying processes will be informed by both previous paradigms,
as our discussion will illustrate.

@@

To emphasize the underlying trend in the changing roles in which we
see and understand noisy processes, I will also restrict my focus 
to relatively simple models primarily from population ecology context.
Simplicity not only makes examples (in equations and in code) more
tractable but also allows us to focus on aspects that are germane to many
contexts rather than unique to particular complexities [@Levins1966; @Bartlett1960].
Nevertheless, that complexity matters -- few themes have been better 
emphasized in the theoretical literature [@Bjornstad2001].  Both
the foundational literature and recent research continue to echo the
theme of understanding the impact different real world complexities
have in stochastic dynamics, including the importance
of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989; @Tuljapurkar1997;
@Caswell2009; @Metcalf2015; @Vindenes2017b], spatial structure [e.g.
@Durrett1994; @Dieckmann2000; @Kerr2002; @Schreiber2010], Individual
heterogeneity [e.g. @Vindenes2015; @Hart2016] or the coexistence
of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007;
@Schreiber2017]; correlated [@Ripa; @Ripa1999; @Petchey; @Schreiber2010;
@Lee2017; @Spanio2017] or periodic [@Coulson2001; @Bjornstad2001;
@Keeling2001] structure in environmental noise, or the interaction
of ecological and evolutionary processes [@Saether1997; @Ozgul2009;
@Lenormand2009; @Vindenes2015; @Schreiber2015].  As such, we will 
rely on both textbooks and recent reviews to provide a proper treatment
of these issues, and focus on broader trends that may be less apparent
to non-specialists.

@@

We choose this model because it has precisely same functional form
as the ubiquitous logistic growth model, $\frac{\ud n}{\ud t} = rn (1 - \tfrac{n}{K})$
but is explicit about how this net growth rate is divided between
birth and death rate.  As we shall see, this is not merely a notational
convenience but leads to real differences in observed stochastic properties
of the model. Both the numerical approach (details in Appendix A) and the 
analytical approach (Appendix B) illustrate that it is straight-forward
to use this same approach in alternate formulations.
Figure \ref{gillespie} shows the results of two exact SSA simulations of this model with
identical parameters except for the total number of available sites,
$N$, illustrating the magnitude of these fluctuations indeed scales in proportion
to $\sqrt{N}$ as postulated by the canonical equation, Eq \eqref{canonical}.

@@

Most ecologists first encounter stochastic models in the context of statistics,
where noise is just a nuisance to be stripped away [@Gotelli2004; @Bolker2007]. 
In examining its origins, we were able to justify the simplistic
view of noise as something merely added on to a deterministic skeleton
to be less ad hoc than it appears; justified as it is by careful
approximations and underlying theorems so long as our systems are 
sufficiently large and the fluctuations sufficiently small.  If this
appears to suggest that noise is somewhat negligible after all as long
as we have the deterministic skeleton, then this section is all about 
rejecting that notion.

@@

Perhaps there is no starker example of differences between the behavior
of a deterministic skeleton and a stochastic model than in the case
of persistence of a species and the coexistence of multiple species.
Demographic noise guarantees extinction in the long-run for any finite
population [e.g. @Kendall1949; @Lande2003;  @Bartlett1955; @Bartlett1960;
@Lande1993; @Saether1998; @Coulson2001; @Schreiber2006;  @Schreiber2010; @Ovaskainen2010;
@Schreiber2017]. Stochasticity can also
do the opposite, allowing many species to coexist in scenarios where
the deterministic skeleton would predict all but one of them to be
doomed to extinction [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Hart2016; @Schreiber2017].
Two important trends in this literature have
been establishing existing results in more general and precise language
and theorems [e.g. see @Schreiber2012 and @Schreiber2017 for an excellent
overview and synthesis of persistence and coexistence respectively],
and continuing to extend such results to more complex contexts 
[@Ovaskainen2010 provides one excellent review], in
particular addressing auto-correlated noise [e.g. @Ripa; @Ripa1999; 
@Petchey; @Schreiber2010; @Lee2017; @Spanio2017] , spatial structure
[e.g. @Durrett1994; @Hastings1994; @Dieckmann2000; @Kerr2002; @Schreiber2010] and stage structure
[e.g. @Cohen1979; @Tuljapurkar1989; @Caswell2009; @Metcalf2015; @Vindenes2017b], 
among other issues. We will turn our attention
away from the well-trodden ground of persistence and coexistence to focus
on other phenomena that can emerge in stochastic models: quasi-cycles and
stochastic switching.  Both examples will illustrate patterns that differ
sharply from those of the deterministic skeleton, both for different underlying
reasons.  Interesting in their own right, these examples will also provide
intuition for our third section on noise the informer.

@@

This review has highlighted contrasting views of the role of noise
between specialists and non-specialists and pointed the way to bridging
that divide by seeing noise not as a new source of information.
Most ecologists are introduced to the concept of stochastic models 
through the lens of statistical inference, in which noise is some
relatively abstract nuisance to be stripped away by clever statistical
inference [@Gotelli2004; @Bolker2007].  As we have seen, specialists have long viewed noise in a 
different light, emerging from the same underlying mechanisms that
create both what we think of as the deterministic processes of interest
and the additional fog of noise [@vanKampen2007; @Black2012].  An intuitive approximation of
larger system size both justifies the convenient partition between
a so-called deterministic skeleton and noise components of a model,
while also illustrating how different demographic and environmental
processes will create different noise dynamics [@Coulson2004; @Ovaskainen2010;
@Schreiber2017].  We have also seen the divide in how noise is perceived
goes well beyond merely a more mechanistic origin story. Specialists
have focused heavily on the role of noise as creator of phenomena:
qualitative differences in the behavior of fully stochastic models and
their deterministic skeletons [@Coulson2004].  Tipping our hat to 
the extensive examples in persistence and coexistence, we explored 
examples of quasi-cycles and stochastic switching showing patterns
wholly unlike their deterministic skeletons. Finally, we endeavor to
bring the divergent views of noise in empirical and theoretical communities
together by turning the noise the nuisance on its head: drawing on
examples from the early warning signals literature which have demonstrated
how noisy phenomena can be used to reveal information about slow
changes in the underlying deterministic skeleton.

Divergent views of the role of stochasticity have separated our community
for too long. The eruption of interest in early warning signals for critical
transitions [@Scheffer2009] provides an excellent illustration of
one way to bridge that divide and the value of greater dialogue between
these two communities.  I hope this review challenges specialists to explore
the potential of stochastic phenomena to inform on underlying processes --
such results are more immediately applicable to empirical work that must
infer process from pattern.  Meanwhile, non-specialists could also
benefit from a more nuanced view of stochasticity as a part of the processes
we seek to understand as well as a novel source of information.  All though
mathematical complexity has frequently been a barrier, computational
advances and the widespread
adoption of computational tools such as the R language have made it possible
to explore these phenomena numerically. Unlike static charts
empirical data, simulations can be manipulated to build
an intuition and generate sample data of these processes. To this end, I 
have provided simple and documented code in Appendix B for each of
examples discussed here.  In the past two
decades, we have set to rest those deterministic skeletons that see
noise only as nuisance and embraced the central role noise plays
as the creator of ecological phenomena. Going forward, I hope and
predict that we will see greater discussion and use of stochastic
phenomena to infer underlying processes in empirical patterns in
similar interactions between theoretical and empirical research.

>>>>

In arguing for this shift, it essential to reconize this is a call for a bigger tent,
not for the rejection of previous paradigms.  What I will characterize as 'noise the nuisance'
reflects a predominately statistical approach, in which noise, almost by definition,
represents all the processes we are not interested in that create additional variation
which might obscure the pattern of interest.  By contrast, an extensive literature has
long explored how noise itself can create patterns and explain processes from population
cycling to coexistence.  These broad categories should be seen as a spectrum and not be
mistaken for either a sharp
dichotomy nor a reference to a stictly empirical-theoretical divide.  Each paradigm
expands upon rather than rejects the previous notion of noise: the recognition that
noise can create novel phenomena does not mean that noise cannot also obscure the
signal of some process of interest.  Likewise, seeking to use noise as a novel source
of information about underlying processes will be informed by both previous paradigms,
as our discussion will illustrate.

@@

To emphasize the underlying trend in the changing roles in which we
see and understand noisy processes, I will also restrict my focus 
to relatively simple models primarily from population ecology context.
Simplicity not only makes examples (in equations and in code) more
tractable but also allows us to focus on aspects that are germane to many
contexts rather than unique to particular complexities [@Levins1966; @Bartlett1960].
Nevertheless, that complexity matters -- few themes have been better 
emphasized in the theoretical literature [@Bjornstad2001].  Both
the foundational literature and recent research continue to echo the
theme of understanding the impact different real world complexities
have in stochastic dynamics, including the importance
of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989; @Tuljapurkar1997;
@Caswell2009; @Metcalf2015; @Vindenes2017b], spatial structure [e.g.
@Durrett1994; @Dieckmann2000; @Kerr2002; @Schreiber2010], Individual
heterogeneity [e.g. @Vindenes2015; @Hart2016] or the coexistence
of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007;
@Schreiber2017]; correlated [@Roughgarden; @Ripa; @Petchey; @Schreiber2010;
@Lee2017; @Spanio2017; @Paniw2018] or periodic [@Coulson2001; @Bjornstad2001;
@Keeling2001] structure in environmental noise, or the interaction
of ecological and evolutionary processes [@Saether1997; @Ozgul2009;
@Lenormand2009; @Vindenes2015; @Schreiber2015].  As such, we will 
rely on both textbooks and recent reviews to provide a proper treatment
of these issues, and focus on broader trends.

@@

We choose this model because it has the same functional form
as the ubiquitous logistic growth model, $\frac{\ud n}{\ud t} = rn (1 - \tfrac{n}{K})$
but is explicit about how this net growth rate is divided between
birth and death rate.  This leads to real differences in observed stochastic properties
of the model. Both the numerical approach (details in Appendix A) and the 
analytical approach (Appendix B) illustrate that it is straight-forward
to use this same approach in alternate formulations.
Figure \ref{gillespie} shows the results of two exact SSA simulations of this model with
identical parameters except for the total number of available sites,
$N$, illustrating the magnitude of these fluctuations indeed scales in proportion
to $\sqrt{N}$ as postulated by the canonical equation, Eq \eqref{canonical}.

@@

Most ecologists first encounter stochastic models in the context of statistics,
where noise is a nuisance to be stripped away [@Gotelli2004; @Bolker2007] to reveal
deterministic mechanisms beneath, rather than being seen as a mechanism in and of itself. 
In examining the origins of noise, we were able to justify the simplistic
view of noise as something merely added on to a deterministic skeleton
to be less ad hoc than it appears; justified as it is by careful
approximations and underlying theorems so long as our systems are 
sufficiently large and the fluctuations sufficiently small.  If this
appears to suggest that noise is somewhat negligible after all as long
as we have the deterministic skeleton, then this section is all about 
rejecting that notion.

@@

Perhaps there is no starker example of differences between the behavior
of a deterministic skeleton and a stochastic model than in the case
of persistence of a species and the coexistence of multiple species.
Demographic noise guarantees extinction in the long-run for any finite
population [e.g. @Kendall1949; @Lande2003;  @Bartlett1955; @Bartlett1960;
@Lande1993; @Saether1998; @Coulson2001; @Schreiber2006;  @Schreiber2010; @Ovaskainen2010;
@Schreiber2017]. Stochasticity can also
do the opposite, allowing many species to coexist in scenarios where
the deterministic skeleton would predict all but one of them to be
doomed to extinction [@Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Hart2016; @Schreiber2017].
Two important trends in this literature have
been establishing existing results in more general and precise language
and theorems [e.g. see @Schreiber2012 and @Schreiber2017 for an excellent
overview and synthesis of persistence and coexistence respectively],
and continuing to extend such results to more complex contexts 
[@Ovaskainen2010 provides one excellent review], in
particular addressing auto-correlated noise [e.g. @Ripa; @Ripa1999; 
@Petchey; @Schreiber2010; @Lee2017; @Spanio2017] , spatial structure
[e.g. @Durrett1994; @Hastings1994; @Dieckmann2000; @Kerr2002; @Schreiber2010] and stage structure
[e.g. @Cohen1979; @Tuljapurkar1989; @Caswell2009; @Metcalf2015; @Vindenes2017b], 
among other issues. We will turn our attention
away from the well-trodden ground of persistence and coexistence to focus
on other phenomena that can emerge in stochastic models: quasi-cycles and
stochastic switching.  Both examples will illustrate patterns that differ
sharply from those of the deterministic skeleton, both for different underlying
reasons.  Interesting in their own right, these examples will also provide
intuition for our third section on noise the informer.

@@

This review has explored three paradigms in how noise is viewed throughout the 
ecological literature, which I have dubbed respectively: noise the *nuisance*, 
noise the *creator*, and noise the *informer*.  Noise can be seen as a nuisance
almost by definition: in examining the origins of noise, we have seen how 
stochasticity is introduced not because ecological processes are random in some fundamental
sense, but rather, because those processes are influenced by a complex combination
of forces we do not model explicitly.  In this view, noise captures all that additional
variation that is separate from the process of interest, and a rich array of 
statistical methods allow us to separate the one from the other in observations and
experiments.  By examining the origins of noise, we have seen that despite the complex
ways in this noise can enter a model, that a Gaussian white-noise approximation [@vanKampen2007; @Black2012] is often
approriate given a limit of a large system size -- a fact often invoked implicitly but
rarely derived explicitly from the theorems of @Kurtz1978 and others.

Building on these foundations, we turned to noise the creator, illustrating how
even small magnitude Gaussian noise could itself create and drive interesting phenomena.
While a purely statistical paradigm might look to explain patterns such as oscillations
or sudden transistions in terms of deterministic processes, this section highlighted how
noise can create and sustain cycles [e.g. @Nisbet1976; @Bjornstad2004] and switches [@Keeling2001].  While our examples focused on the most
tractable systems, a wealth of literature has explored such phenomena in ever more complex
contexts.  These examples paint a very different picture of noise, one where "everything matters" [@Bjornstad2001],
where it can be difficult to know what drives a pattern and where ommitting any of the
complexity (age or spatial structure, autocorrelation, individual heterogeneity) can
qualitatively alter the behavior of a model.

Our third paradigm seeks a more optimistic middle ground of noise the informer.
Here we saw the examples from an empricially driven literature on early warning signals
[@Scheffer2009; @Dai2012] view noise as a source of countless miniature experiments
which can reveal the underlying dynamics of a system and how they may be slowly changing.
In this context, noise does not act to create phenomena of interest directly. The 
sudden transistions we seek to anticipate are still explained by the deterministic part
of the model -- bifurcations. But nor is noise a nuisance that merely cloaks this
deterministic skeleton from plain view: rather, it becomes a novel source of information
that would be inaccessible from a purely deterministic approach.  I believe more examples
of how noise can inform on underlying processes is possible, but will require greater
dialog between these world views.
